Enabling Ambient Backscatter \\Using a Low-Cost Software Defined Radio

### 2017-04-07 paper.tex
Compressed the receiver part once more.

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Enabling Ambient Backscatter \\Using Low-Cost Software Defined Radio}

@@

A backscattering antenna is the most important part of backscatter
transmitter. Our antenna is optimized for usage with the television
signal, which has its center frequency at 626 MHz. It is a ground plane
antenna, which has been built up on a large unetched PCB plate. A hole
in the middle of the PCB serves for connecting a wire to a 50
\ensuremath{\Omega} micro strip transmission line. The wire, which acts
as a vertical antenna element is soldered to the transmission line. The
transmission line has been calculated for 626 MHz using an FR4 substrate
with a thickness of 1.6 mm. The design equations given in
\cite{pozar_microwave_2011} are implemented in various programs e.g.
KiCAD, which has been used for solving the design equations. An
open-stub, attached directly at the juncture where the vertical antenna
element is attached to the transmission line, is used for fine tuning of
the input impedance (cf. chapter 5 in \cite{pozar_microwave_2011}).
Accepting -15 dB as sufficient value for \ensuremath{S_{11}}, one can
say, that the antenna works in a range from 626 MHz to 645 MHz. For
backscattering this small bandwidth is not necessarily a disadvantage,
since everything that is within the operating frequency of the antenna
is backscattered. With our narrowband antenna the undesired
backscattering of out-of-band signals is reduced.  With our approach we
were able to achieve an input impedance at 634 MHz of
\ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz
\ensuremath{S_{\text{11}}} is still approximately -17 dB.

@@

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
C++ has been used to implement the receiver. We provide the code
publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}. A highly
sophisticated bus system has been developed to exchange data between the
different components of the system. The signal flow in figure
\ref{fig:receiver_arch} is from the left to the right.
\textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on
\textit{libusb}, is used to transfer the data from the TV stick into our
program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes
down the high frequency to a intermediate frequency (both values are
software adjustable). Subsequently the data is mixed with sine and
cosine, low-pass filtered and finally transferred to the digital domain.
This data is available as two 8-bit values (real and imaginary). With
this two values, the phase and absolute value of the data can be
determined for every sample. It can be intepreted can be seen as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

So the first block, which is entitled RTL2832U,
provides the interface to the TV dongle. This block is represented as
one object. It can set the frequency \ensuremath{f_{tuned}}, where the
receiver is listening, besides other interesting parameters e.g. the
analog gain.
The demodulator block is responsible for converting the sampled signals
into ones and zeroes. Therefore the real and imaginary values first have
to be deinterleaved out of the sample message. Now they are represented
as integer values between 0 and 255 (8 bit). So one converts them to
floats, whereas 127.5 equals 0. After that the values have to be
downsampled. Usually we were sampling with 250 kHz. This is still much
more than we need assuming an Additive White Gaussian Noise (AWGN)
channel with a \ensuremath{S/N = 10\,dB} (cf.  equation \ref{eq:awgn}).
So we decided to reduce the sampling frequency to 25 kHz.
\begin{equation}
	\label{eq:awgn}
	C=B \cdot log_2(1+\frac{S}{N})= 25\,kHz \cdot log_2(11) \approx 86.5\,kbit/s
\end{equation}     
To do this it is important to use an anti-aliasing filter, as the Shannon-Nyquist theorem has to be satisfied: 
\begin{equation}
	\label{eq:nyquist}
	f_{samp} \geq 2 \cdot f_{a}
\end{equation}
\ensuremath{f_{a}} is the highest frequency in the signal. 
 A FIR
filter naively implemented with floats has been used to achieve this. 10
kHz is the cut-off frequency of the first filter. After downsampling
another filter is used to suppress noise. This filter has a cut-off
frequency of 3 kHz. The last demodulation step is to rectify the signal
(cf. equation \ref{eq:rectify}) and decide with a software-defined
Schmitt trigger whether a sample is a 1 or a 0.  
\begin{equation}
	\label{eq:rectify}
	abs(I,Q) = \sqrt{I^2 +  Q^2}
\end{equation} 
These samples are broadcast on the bus. A registered listener receives
the processed samples. This registered listener is the decoder, which is
represented as an object as well. After the decoder received the samples
it decides when a frame is sent or when the channel is idle.  The
decoder also includes a function to correlate the received bitstream
with an expected pattern. Hence it is able to determine the bit error
ratio (BER). Furthermore the code we have written so far subsumes
different simulators for e.g. playing back recoreded data. Another handy
utility, which has been written in Octave, is a tool (an oscilloscope)
to look at the data, which is written into files by the demodulator. An
example of a recorded data stream with 25 kS/s can be seen in figure
\ref{fig:transmission}.  \begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength.}
\label{fig:transmission}
\end{figure}

@@

We have used the textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).

@@

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T is specified up to 10 MHz bandwidth.
And the local TV signal provider is to its own statements using DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\bigl( \Re\{ s_{\text{band,n}}(t) \} \bigr) \Biggr|
\end{equation}     
where \ensuremath{N} is the total number of intervals accumulated over
the entire frequency sweep. The FFT
is only carried out over the real values, which are received from the
\textit{RTL-SDR}. \ensuremath{s_{\text{band,n}}(t)} is the signal in the
time domain of one interval (one subband). There are as many samples taken in one band as
needed for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to normalize everything and calculate the power. Before doing that a simple smoothing algorithm (FIR) is applied on \ensuremath{|H|}.
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain
point in space. 
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\caption{Left: Signal strength fading against distance relative
to the spot where the maximum signal strength has been captured (TV
towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been
calculated with the Haversine formula. Right: Heatmap of the 626 MHz TV
signal in Uppsala.}

>>>>

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Enabling Ambient Backscatter \\Using a Low-Cost Software Defined Radio}

@@

Our backscattering antenna is optimized for usage with the television signal,
which has its center frequency at 626 MHz. It is a ground plane antenna, which
has been built up on a large unetched PCB plate. A hole in the middle of the
PCB serves for connecting a wire to a 50 \ensuremath{\Omega} micro strip
transmission line. The wire, which acts as a vertical antenna element is
soldered to the transmission line. The microstrip transmission line design
equations given in \cite{pozar_microwave_2011} are implemented in various
programs e.g.  KiCAD, which has been used for designing them. An open-stub,
attached directly at the juncture where the vertical antenna element is
attached to the transmission line, is used for fine tuning of the input
impedance (cf. chapter 5 in \cite{pozar_microwave_2011}).  Accepting -15 dB as
sufficient value for \ensuremath{S_{11}}, one can say, that the antenna works
within a range from 626 MHz to 645 MHz. With our narrowband antenna the
undesired backscattering of out-of-band signals is reduced. With our approach
we were able to achieve an input impedance at 634 MHz of
\ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz
\ensuremath{S_{\text{11}}} is still approximately -17 dB.

@@

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
We provide our C++ receiver code publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}. A highly sophisticated
bus system has been developed to exchange data between the different components
of the system. The signal flow in figure \ref{fig:receiver_arch} is from the
left to the right.  \textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is
based on \textit{libusb}, is used to transfer the data from the TV stick into
our program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes down
the high frequency to a intermediate frequency (both values are software
adjustable). The sampled data is available as two 8-bit values (real and
imaginary). With this two values, the phase and absolute value of the data can
be determined for every sample. Every sample can be intepreted as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

So the first block, which is entitled RTL2832U, provides the interface to the
TV dongle. This block controls the frequency \ensuremath{f_{tuned}}, where the
receiver is listening, besides other interesting parameters e.g. the analog
gain.

The demodulator block is responsible for converting the sampled signals into
ones and zeroes. Therefore the real and imaginary values first have to be
deinterleaved out of the sample message. These 8 bit integers are converted
into floats, whereas 127.5 equals 0. After that the values have to be
downsampled. Usually we were sampling with 250 kHz. This is still much more
than we need assuming an Additive White Gaussian Noise (AWGN) channel with a
\ensuremath{S/N = 10\,dB} (cf.  equation \ref{eq:awgn}).  So we decided to
reduce the sampling frequency to 25 kHz.

\begin{equation}
	\label{eq:awgn}
	C=B \cdot log_2(1+\frac{S}{N})= 25\,kHz \cdot log_2(11) \approx 86.5\,kbit/s
\end{equation}     
To do this it is important to use an anti-aliasing filter, as the Shannon-Nyquist theorem has to be satisfied: 
\begin{equation}
	\label{eq:nyquist}
	f_{samp} \geq 2 \cdot f_{a}
\end{equation}
\ensuremath{f_{a}} is the highest frequency in the signal. 
 A FIR filter
naively implemented with floats has been used to achieve this.  After
downsampling another filter is used to suppress noise.  The last demodulation
step is to rectify the signal (cf. equation \ref{eq:rectify}) and decide with a
software-defined Schmitt trigger whether a sample is a 1 or a 0.  
\begin{equation}
	\label{eq:rectify}
	abs(I,Q) = \sqrt{I^2 +  Q^2}
\end{equation} 
A registered listener receives the processed samples, which are broadcasted by
the demodulator. This registered listener is the decoder. After the decoder
received the samples it decides when a frame is sent or when the channel is
idle.  The decoder also includes a function to correlate the received bitstream
with an expected pattern. Hence it is able to determine the bit error ratio
(BER). Furthermore the code we have written so far subsumes different
simulators for e.g. playing back recoreded data. Another handy utility, which
has been written in Octave, is a tool (an oscilloscope) to look at the data,
which is written into files by the demodulator (cf. figure \ref{fig:transmission}).
\begin{figure}[h]

\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength. Sampling frequency is 25 kHz.}
\label{fig:transmission}
\end{figure}

@@

We have used the \textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).

@@

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T is specified up to 10 MHz bandwidth.
And the local TV signal provider is to its own statements using DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\biggl( \Re\{ s_{\text{band,n}}(t) \} \biggr) \Biggr|
\end{equation}     
where \ensuremath{N} is the total number of intervals accumulated over
the entire frequency sweep. The FFT
is only carried out over the real values, which are received from the
\textit{RTL-SDR}. \ensuremath{s_{\text{band,n}}(t)} is the signal in the
time domain of one interval (one subband). There are as many samples taken in one band as
needed for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to normalize everything and calculate the power. Before doing that a simple smoothing algorithm (FIR) is applied on \ensuremath{|H|}.
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain
point in space. 
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\caption{Left: Signal strength fading against distance relative
to the spot where the maximum signal strength has been captured (TV
towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been
calculated with the Haversine formula. Right: Heatmap of the 626 MHz TV
signal in Uppsala.}
